On the Design, Implementation and Use of Laziness in R

### 2019-08-03 report/garbage.Rmd
Add promise gc analysis and parameterize report

```{r echo = FALSE, warning=FALSE, message=FALSE}
Sys.setenv("DISPLAY"=":0")

@@

function_call_summary <- read_table("function_call_summary")
grouped_object_count_by_type <- read_table("grouped_object_count_by_type")
argument_promise_count_by_value_type <- read_table("argument_promise_count_by_value_type")
argument_promise_count_by_expression_type <- read_table("argument_promise_count_by_expression_type")
promise_count_by_category <- read_table("promise_count_by_category")
argument_promise_count_by_dispatch_type <- read_table("argument_promise_count_by_dispatch_type")
argument_promise_count_by_force_count <- read_table("argument_promise_count_by_force_count")
package_strictness <- read_table("package_strictness")
formal_parameter_count_by_use <- read_table("formal_parameter_count_by_use")
argument_nature <- read_table("promise_argument_count_by_nature")
formal_parameter_count_by_usage_class <- read_table("formal_parameter_count_by_usage_class")
argument_promise_count_by_value_lookup_count <- read_table("argument_promise_count_by_value_lookup_count")
execution_times <- read_table("execution_times")
call_by_escape_and_type <- read_table("function_call_count_by_escape_category_and_return_value_type")
escaped_argument_count_by_nature <- read_table("escaped_argument_count_by_nature")
argument_promise_use_distribution <- read_table("argument_promise_use_distribution")
package_metaprogramming <- read_table("package_metaprogramming")
argument_expression_type_by_usage_class <- read_table("argument_expression_type_by_usage_class")
force_orders <- read_table("closure_count_by_force_order_count")
extreme_event_counts <- read_table("extreme_event_counts")
event_processing_rates <- read_table("event_processing_rates")
argument_type <- read_table("argument_count_by_type")
closure_count_by_call_count <- read_table("closure_count_by_call_count")
```

# Overview

## Functions and Calls
```{r, eval = TRUE, echo = FALSE, results = "hide"}
delayed_assign_id <- "zU2zBIO6UWH#zA6LsJRrFw=="
force_id <- "K0njDVAdxhn0LsQoHtOLWA=="
force_and_call_id <- "forceAndCall"
substitute_id <- "substitute"
eval_id <- "d7Cor#TdZ2FDZivyZksQqQ=="
evalq_id <- "JYSerovrw0rT49mweaQtQw=="
eval_parent_id <- "JRhrApn8kt7E6bPe#W0wYA=="
local_id <- "9ltwkYFALCIgkTUFD7LNLw=="

total_call_count <- sum(function_call_summary$call_count)

total_function_count <- length(unique(function_call_summary$function_id))

get_call_count <- function(df, id) {
    id <- enquo(id)
    function_call_summary %>%
        filter(function_id == !!id) %>%
        pull(call_count) %>%
        as.double() %>%
        sum()
}

corpus_counts <- 
    tibble(name = character(0), value = double(0)) %>%
    add_row(name = "CountCall", value = total_call_count) %>%
    add_row(name = "CountFunction", value = total_function_count)

jumped_call_count <-
    function_call_summary %>%
    filter(jumped) %>%
    pull(call_count) %>%
    sum()

corpus_counts <-
    corpus_counts %>%
    add_row(name = "CountJumpedCall", value = jumped_call_count) %>%
    add_row(name = "PercJumpedCall", value = jumped_call_count / total_call_count)

unjumped_call_count <-
    function_call_summary %>%
    filter(!jumped) %>%
    pull(call_count) %>%
    sum()

corpus_counts <-
    corpus_counts %>%
    add_row(name = "CountUnjumpedCall", value = unjumped_call_count) %>%
    add_row(name = "PercUnjumpedCall", value = unjumped_call_count / total_call_count)

function_call_summary <-
    function_call_summary %>%
    filter(!jumped) %>%
    select(-jumped)

unjumped_function_count <-
    function_call_summary %>%
    pull(function_id) %>%
    unique() %>%
    length()

corpus_counts <-
    corpus_counts %>%
    add_row(name = "CountUnjumpedFunction", value = unjumped_function_count) %>%
    add_row(name = "PercUnjumpedFunction", value = unjumped_function_count / total_function_count)

builtin_call_count <-
    function_call_summary %>%
    filter(function_type == "Builtin") %>%
    pull(call_count) %>%
    sum()

corpus_counts <-
    corpus_counts %>%
    add_row(name = "CountBuiltinCall", value = builtin_call_count) %>%
    add_row(name = "PercBuiltinCall", value = builtin_call_count / unjumped_call_count)

special_call_count <-
    function_call_summary %>%
    filter(function_type == "Special") %>%
    pull(call_count) %>%
    sum()

corpus_counts <-
    corpus_counts %>%
    add_row(name = "CountSpecialCall", value = special_call_count) %>%
    add_row(name = "PercSpecialCall", value = special_call_count / unjumped_call_count)

closure_call_count <-
    function_call_summary %>%
    filter(function_type == "Closure") %>%
    pull(call_count) %>%
    sum()

corpus_counts <-
    corpus_counts %>%
    add_row(name = "CountClosureCall", value = closure_call_count) %>%
    add_row(name = "PercClosureCall", value = closure_call_count / unjumped_call_count)

builtin_function_count <-
    function_call_summary %>%
    filter(function_type == "Builtin") %>%
    pull(function_id) %>%
    unique() %>%
    length()

corpus_counts <-
    corpus_counts %>%
    add_row(name = "CountBuiltinFunction", value = builtin_function_count)

special_function_count <-
    function_call_summary %>%
    filter(function_type == "Special") %>%
    pull(function_id) %>%
    unique() %>%
    length()

corpus_counts <-
    corpus_counts %>%
    add_row(name = "CountSpecialFunction", value = special_function_count)

closure_function_count <-
    function_call_summary %>%
    filter(function_type == "Closure") %>%
    pull(function_id) %>%
    unique() %>%
    length()

corpus_counts <-
    corpus_counts %>%
    add_row(name = "CountClosureFunction", value = closure_function_count)

s3_function_count <-
    function_call_summary %>%
    filter(function_type == "Closure") %>%
    filter(S3_call_count > 0) %>%
    pull(function_id) %>%
    unique() %>%
    length()

corpus_counts <-
    corpus_counts %>%
    add_row(name = "CountThreeFunction", value = s3_function_count) %>%
    add_row(name = "PercThreeFunction", value = s3_function_count / closure_function_count)

s3_call_count <-
    function_call_summary %>%
    filter(function_type == "Closure") %>%
    pull(S3_call_count) %>%
    sum()

corpus_counts <-
    corpus_counts %>%
    add_row(name = "CountThreeCall", value = s3_call_count) %>%
    add_row(name = "PercThreeCall", value = s3_call_count / closure_call_count)

s4_function_count <-
    function_call_summary %>%
    filter(function_type == "Closure") %>%
    filter(S4_call_count > 0) %>%
    pull(function_id) %>%
    unique() %>%
    length()

corpus_counts <-
    corpus_counts %>%
    add_row(name = "CountFourFunction", value = s4_function_count) %>%
    add_row(name = "PercFourFunction", value = s4_function_count / closure_function_count)

s4_call_count <-
    function_call_summary %>%
    filter(function_type == "Closure") %>%
    pull(S4_call_count) %>%
    sum()

corpus_counts <-
    corpus_counts %>%
    add_row(name = "CountFourCall", value = s4_call_count) %>%
    add_row(name = "PercFourCall", value = s4_call_count / closure_call_count)

ordinary_function_count <-
    function_call_summary %>%
    filter(function_type == "Closure") %>%
    filter(S4_call_count == 0 & S3_call_count == 0) %>%
    pull(function_id) %>%
    unique() %>%
    length()

corpus_counts <-
    corpus_counts %>%
    add_row(name = "CountOrdFunction", value = ordinary_function_count) %>%
    add_row(name = "PercOrdFunction", value = ordinary_function_count / closure_function_count)

ordinary_call_count <-
    function_call_summary %>%
    filter(function_type == "Closure") %>%
    mutate(ordinary_call_count = call_count - S3_call_count - S4_call_count) %>%
    pull(ordinary_call_count) %>%
    sum()

corpus_counts <-
    corpus_counts %>%
    add_row(name = "CountOrdCall", value = ordinary_call_count) %>%
    add_row(name = "PercOrdCall", value = ordinary_call_count / closure_call_count)

specific_counts <-
    tibble(name = c("DelayedAssign", "Force", "ForceAndCall", "Substitute", "Eval", "EvalQ", "EvalParent", "Local"),
           id = c(delayed_assign_id, force_id, force_and_call_id, substitute_id, eval_id,  evalq_id,  eval_parent_id, local_id)) %>%
    mutate(call_count = map_dbl(id, function(id) get_call_count(function_call_summary, id)))

specific_percs <-
    specific_counts %>%
    mutate(relative_call_count = call_count / closure_call_count)

corpus_counts <-
    corpus_counts %>%
    rbind(
        specific_counts %>%
        mutate(name = str_c("Count", name, "Call")) %>%
        select(name, value = call_count)
    ) %>%
    rbind(
        specific_percs %>%
        mutate(name = str_c("Perc", name, "Call")) %>%
        select(name, value = relative_call_count)
    )

total_parameter_count <- sum(as.double(formal_parameter_count_by_use$parameter_count))

total_argument_count <- sum(as.double(argument_nature$argument_count))

default_argument_count <-
    argument_nature %>%
    filter(argument_nature == "Default") %>%
    pull(argument_count)

non_default_argument_count <-
    argument_nature %>%
    filter(argument_nature == "Non Default") %>%
    pull(argument_count)

corpus_counts <-
    corpus_counts %>%
    add_row(name = "CountParameter", value = total_parameter_count) %>%
    add_row(name = "CountArgument",  value = total_argument_count) %>%
    add_row(name = "CountDefArgument",  value = default_argument_count) %>%
    add_row(name = "PercDefArgument",  value = default_argument_count / total_argument_count) %>%
    add_row(name = "CountNonDefArgument",  value = non_default_argument_count) %>%
    add_row(name = "PercNonDefArgument",  value = non_default_argument_count / total_argument_count)

missing_argument_count <- 
    argument_type %>%
    filter(argument_type == "Missing") %>%
    pull(argument_count)

relative_missing_argument_count <- 
    argument_type %>%
    filter(argument_type == "Missing") %>%
    pull(relative_argument_count)

promise_argument_count <-
    argument_type %>%
    filter(argument_type == "Promise") %>%
    pull(argument_count)

relative_promise_argument_count <- 
    argument_type %>%
    filter(argument_type == "Promise") %>%
    pull(relative_argument_count)

corpus_counts <-
    corpus_counts %>%
    add_row(name = "CountMissingArgument", value = missing_argument_count) %>%
    add_row(name = "PercMissingArgument", value = relative_missing_argument_count) %>%
    add_row(name = "CountPromiseArgument", value = promise_argument_count) %>%
    add_row(name = "PercPromiseArgument", value = relative_promise_argument_count)

event_processing_rates <-
    event_processing_rates %>%
    filter(tracing_time > 0)

avg_event_processing_rate <-

avg_event_processing_rate <- mean(event_processing_rates$event_rate)
avg_expression_processing_rate <- mean(event_processing_rates$expression_rate)

event_processing_rate <- 
    sum(as.double(event_processing_rates$event_count)) / sum(event_processing_rates$tracing_time)
expression_processing_rate <-
    sum(as.double(event_processing_rates$expression_count)) / sum(event_processing_rates$tracing_time)

corpus_counts <-
    corpus_counts %>%
    add_row(name = "CountEvent", value = sum(as.double(event_processing_rates$event_count))) %>%
    add_row(name = "CountTotEventRate", value = event_processing_rate) %>%
    add_row(name = "CountAvgEventRate", value = avg_event_processing_rate) %>%
    add_row(name = "CountExpression", value = sum(as.double(event_processing_rates$expression_count))) %>%
    add_row(name = "CountTotExpressionRate", value = expression_processing_rate) %>%
    add_row(name = "CountAvgExpressionRate", value = avg_expression_processing_rate)

e <- environment()

corpus_counts %>%
    pmap(function(name, value) {
        generate_macro(name, value, subst = FALSE)
        assign(name, value, envir = e)
    })
```

We observe `r format_as_absolute(CountCall)` calls to `r format_as_absolute(CountFunction)` functions. 
Out of these, `r format_as_absolute(CountJumpedCall)` calls exit early due to longjump (exception handling or non-local promise returns).
Of the remaining `r format_as_absolute(CountUnjumpedCall)` (`r format_as_percentage(PercUnjumpedCall)`) calls to 
`r format_as_absolute(CountUnjumpedFunction)` (`r format_as_percentage(PercUnjumpedFunction)`) functions, 
`r format_as_absolute(CountBuiltinCall)` (`r format_as_percentage(PercBuiltinCall)`) calls are made to 
`r format_as_absolute(CountBuiltinFunction)` builtin functions, 
`r format_as_absolute(CountSpecialCall)` (`r format_as_percentage(PercSpecialCall)`) calls are made to 
`r format_as_absolute(CountSpecialFunction)` special functions and 
`r format_as_absolute(CountClosureCall)` (`r format_as_percentage(PercClosureCall)`) calls are made to 
`r format_as_absolute(CountClosureFunction)` R functions. 
Of these calls to R functions 
`r format_as_absolute(CountThreeCall)` (`r format_as_percentage(PercThreeCall)`) calls are made to 
`r format_as_absolute(CountThreeFunction)` (`r format_as_percentage(PercThreeFunction)`) S3 methods,
`r format_as_absolute(CountFourCall)` (`r format_as_percentage(PercFourCall)`) calls are made to 
`r format_as_absolute(CountFourFunction)` (`r format_as_percentage(PercFourFunction)`) S4 methods and 
`r format_as_absolute(CountOrdCall)` (`r format_as_percentage(PercOrdCall)`) calls are made to 
`r format_as_absolute(CountOrdFunction)` (`r format_as_percentage(PercOrdFunction)`) ordinary R functions.
`r format_as_absolute(CountDelayedAssignCall)` (`r format_as_percentage(PercDelayedAssignCall)`) calls are made to the `delayedAssign` function, 
`r format_as_absolute(CountForceCall)` (`r format_as_percentage(PercForceCall)`) calls are made to the `force` function, 
`r format_as_absolute(CountForceAndCallCall)` (`r format_as_percentage(PercForceAndCallCall)`) calls are made to `forceAndCall` function and 
`r format_as_absolute(CountSubstituteCall)` (`r format_as_percentage(PercSubstituteCall)`) calls to the `substitute` function.

The following graph shows the distribution of packages by the number of functions observed in them.

```{r, eval = TRUE, echo = FALSE}
total_package_count <- nrow(package_strictness)
### geom_histogram did not let me add the other y axis in a reasonable way,
### so, I explicitly computed the histogram and used geom_col instead.
package_function_distribution <-
    package_strictness %>%
    filter(type == "compatible") %>%
    group_by(total_function_count) %>%
    summarize(relative_package_count = n() / total_package_count) %>%
    ungroup()

dense_package_count <-
    package_function_distribution %>%
    filter(total_function_count > 50) %>%
    pull(relative_package_count) %>%
    sum()

package_function_distribution <-
    package_function_distribution %>%
    filter(total_function_count <= 50) %>%
    mutate(total_function_count = as.character(total_function_count)) %>%
    add_row(total_function_count = "> 50", relative_package_count = dense_package_count)

package_function_graph <-
    package_function_distribution %>%
    ggplot(aes(x = total_function_count, y = relative_package_count)) +
    geom_col() + 
    #coord_flip() +
    scale_x_discrete(limits = c(1:50, "> 50")) +
    scale_y_continuous(sec.axis = sec_axis(~ .* total_package_count, labels=count_labels),
                       labels = relative_labels) +
    labs(x = "", y = "")

save_local_graph(package_function_graph, "package_function.pdf", width = 12, height = 3)
```

The following graph shows the distribution of R functions by the number of calls.

```{r, eval = TRUE, echo = FALSE}
total_closure_count <- 
    function_call_summary %>%
    filter(function_type == "Closure") %>%
    pull(function_id) %>%
    unique() %>%
    length()

closure_function_count <-
    function_call_summary %>%
    filter(function_type == "Closure") %>%
    pull(function_id) %>%
    unique() %>%
    length()

closure_call_distribution_graph <-
    closure_count_by_call_count %>%
    filter(call_count %in%  c(1:10, "> 10")) %>%
    ggplot(aes(x = call_count,
               y = relative_closure_count)) +
    geom_col() +
    scale_x_discrete(limits = c(1:10, "> 10")) +
    scale_y_continuous(sec.axis = sec_axis(~ . * total_closure_count, labels = count_labels),
                       labels = relative_labels) +
    labs(x = "", y = "") +
    coord_flip()

save_local_graph(closure_call_distribution_graph, "closure_call_distribution_graph.pdf", width = 12, height = 3)
```

We observe that most functions are called more than 10 times.

## Parameters and Arguments

We observed `r format_as_absolute(CountParameter)` formal parameter positions.
Of the `r format_as_absolute(CountArgument)` arguments `r format_as_absolute(CountDefArgument)` (`r format_as_percentage(PercDefArgument)`) were default arguments and the remaining `r format_as_absolute(CountNonDefArgument)` (`r format_as_percentage(PercNonDefArgument)`) were non-default arguments. `r format_as_absolute(CountMissingArgument)` (`r format_as_percentage(PercMissingArgument)`) arguments were missing and `r format_as_absolute(CountPromiseArgument)` were promises.

The following graph shows the distribution of R functions by their formal parameter counts.

```{r, eval = TRUE, echo = FALSE}
total_closure_count <-
    function_call_summary %>%
    filter(function_type == "Closure") %>%
    pull(function_id) %>%
    unique() %>%
    length()

closure_count_by_formal_parameter_count <-
    function_call_summary %>%
    filter(function_type == "Closure") %>%
    group_by(formal_parameter_count) %>%
    summarize(closure_count = length(unique(function_id))) %>%
    ungroup() %>%
    mutate(relative_closure_count = closure_count / sum(closure_count)) %>%
    summarize_outliers(formal_parameter_count, closure_count, relative_closure_count, 5)

closure_parameter_distribution_graph <-
    closure_count_by_formal_parameter_count %>%
    filter(formal_parameter_count %in% c(0:5, "> 5")) %>%
    ggplot(aes(x = formal_parameter_count,
               y = relative_closure_count)) +
    geom_col() +
    scale_x_discrete(limits = c(0:5, "> 5")) +
    scale_y_continuous(sec.axis = sec_axis(~ . * total_closure_count, labels = count_labels),
                       labels = relative_labels) +
    labs(x = "", y = "") +
    coord_flip()

save_local_graph(closure_parameter_distribution_graph, "closure_parameter_distribution_graph.pdf", width = 12, height = 3)
```

## Events and Expressions

`r format_as_absolute(CountEvent)` events were analyzed from `r format_as_absolute(CountExpression)` expressions.
The total event processing rate was `r format_as_absolute(CountTotEventRate)` events per second.
The total expression processing rate was `r format_as_absolute(CountTotExpressionRate)` expression evaluations per second.

# Object Counts (Fig. 3)

The following graph  shows the distribution of object counts by type.

```{r, echo = FALSE, eval = TRUE}
total_object_count <-
    grouped_object_count_by_type %>%
    pull(count) %>%
    sum()

## Let's get rid of funny types
grouped_object_count_by_type <-
   grouped_object_count_by_type %>%
   mutate(group_type = ifelse(type=="S4","S4",group_type)) %>%
   mutate(group_type = ifelse(type=="Symbol","Symbol",group_type)) %>%
   filter(group_type != "Other")

object_count_graph <-
    grouped_object_count_by_type %>%
    mutate(relative_count = count/total_object_count) %>%
    ggplot(aes(x = reorder(group_type, -count),
               y = relative_count, fill = type)) +
    geom_col(fill="grey45") + 
    coord_flip() +
    scale_y_continuous(sec.axis = sec_axis(~ .*total_object_count, labels=count_labels),
                       labels = relative_labels) +
    labs(x = "", y = "") +
    theme(legend.title = element_blank(), legend.position = "top")

save_local_graph(object_count_graph, "grouped_object_count_by_type.pdf", width = 12, height = 4)
```

```{r, eval = TRUE, echo = FALSE}
CountProm <- sum(promise_count_by_category$promise_count)
generate_macro("CountProm", CountProm)

PercArgProm <- promise_count_by_category$relative_promise_count[1]
generate_macro("PercArgProm", PercArgProm)

PercForceProm <- argument_promise_count_by_force_count$relative_promise_count[2]
generate_macro("PercForceProm", PercForceProm)

CountForceProm <-
    argument_promise_count_by_force_count %>% 
    filter(force_count>1) %>% 
    pull(promise_count) %>% 
    sum()
generate_macro("CountForceProm", CountForceProm)

PercDispatch <- 
    argument_promise_count_by_dispatch_type %>% 
    filter(dispatch_type != "None") %>% 
    pull(relative_promise_count) %>% 
    sum()
generate_macro("PercDispatch", PercDispatch)
```

We observed `r format_as_absolute(CountProm)` promises, out of which `r format_as_percentage(PercArgProm)` were created as part of argument lists. `r format_as_percentage(PercForceProm)` of promises were forced. `r format_as_percentage(PercDispatch)` of promises were forced due to dispatch.

# Promise Result Types (Fig. 4)

The figure below shows the content of forced promises.
The most common types are character strings, logicals, environments, double, integer and closure. The
presence of null values is explained by the fact that many default argument values are set to null.

```{r, eval = TRUE, echo = FALSE}
argument_promise_count_by_value_type <- 
    argument_promise_count_by_value_type %>%
    rename(type = value_type, count = promise_count)

argument_promise_count_by_value_type <-
    argument_promise_count_by_value_type %>%
    filter(type!="Unbound" ,type!="Function Call", type!="...", type!="Builtin", type!="Bytecode") %>%
    filter(type!="Expression", type!="Externalptr", type!="Special", type!="Unknown (Jumped)") %>%
    filter(type!="Pairlist", type!="Missing", type!="Weakref")

total_count <-  sum(argument_promise_count_by_value_type$count)

argument_promise_count_by_value_type_graph <-
    argument_promise_count_by_value_type %>%
    ggplot(aes(reorder(type, -count), relative_promise_count)) +
    geom_col(fill="grey45") + 
    coord_flip() +
    scale_y_continuous(sec.axis = sec_axis(~ . * total_count, labels = count_labels),
                       labels = relative_labels) +
    labs(x = "", y = "") +
    theme(legend.title = element_blank(), legend.position = "top")

save_local_graph(argument_promise_count_by_value_type_graph, "prom_types.pdf", width = 12, height = 4)
```

# Promise Expression Types (Fig. 5)

Figure below shows the content of the expression slot of promises. The majority contain a single symbol to be looked
up, e.g. x. Of course that symbol may be bound to another promise. There are some function calls, e.g. `1 + 2`. 
The remainder are inlined values (scalar constants).

```{r, eval = TRUE, echo = FALSE}
argument_promise_count_by_expression_type <- 
    argument_promise_count_by_expression_type %>% 
    rename(type=expression_type, count=promise_count)

argument_promise_count_by_expression_type <-
    argument_promise_count_by_expression_type %>% 
    filter(type!="Unbound", type!="...",type!="Builtin",type!="Bytecode") %>%
    filter(type!="Expression",type!="Externalptr",type!="Special",type!="Unknown (Jumped)") %>%
    filter(type!="Pairlist",type!="Missing",type!="Weakref",type!="Promise")

total_count <-  sum(argument_promise_count_by_expression_type$count)

argument_promise_count_by_expression_type_graph <-
    argument_promise_count_by_expression_type %>%
    ggplot(aes(reorder(type, -count),
               relative_promise_count)) +
    geom_col(fill="grey45") + coord_flip() +
    scale_y_continuous(sec.axis = sec_axis(~ . * total_count, labels = count_labels),
                       labels = relative_labels) +
    labs(x = "", y = "") +
    theme(legend.title = element_blank(),
          legend.position = "top")

save_local_graph(argument_promise_count_by_expression_type_graph, "prom_exp_types.pdf", width = 12, height = 4)
```

# Function Strictness per package (Fig. 6)

```{r, eval = TRUE, echo=FALSE}
package_strictness <- 
   package_strictness %>%
   mutate(package = package_name) %>%
   filter(!(package_name %in% c("global", "empty"))) %>%
   filter(type == "compatible")

tot_strict <-
    package_strictness %>%
    pull(strict_function_count) %>%
    sum()

CountFun <-
    package_strictness %>%
    pull(total_function_count) %>%
    sum()

PercStrictFun <- tot_strict / CountFun

generate_macro("PercStrictFun", PercStrictFun)

generate_macro("CountFun", CountFun)

CountStrFunLess <-
    package_strictness %>%
    filter(relative_strict_function_count < .75) %>% 
    pull(total_function_count) %>% 
    sum()

CountStrPackLess <-
    package_strictness %>%
    filter(relative_strict_function_count <.75) %>% 
    nrow()

CountStrPackMore <-
    package_strictness %>%
    filter(relative_strict_function_count >.75) %>% 
    pull(total_function_count) %>% 
    sum()

CountStrFunMore <-
    package_strictness %>%
    filter( relative_strict_function_count >.75) %>% 
    nrow()

PercStrPackLess <- CountStrPackLess/nrow(package_strictness)

PercStrFunLess <- CountStrFunLess/sum(package_strictness$total_function_count)

generate_macro("CountStrPackLess", CountStrPackLess)
generate_macro("CountStrFunLess", CountStrFunLess)
generate_macro("PercStrPackLess", PercStrPackLess)
generate_macro("PercStrFunLess", PercStrFunLess)
generate_macro("CountStrPackMore", CountStrPackMore)
generate_macro("CountStrFunMore", CountStrFunMore)

total_package_count <- nrow(package_strictness)
### geom_histogram did not let me add the other y axis in a reasonable way,
### so, I explicitly computed the histogram and used geom_col instead.
package_strictness_graph <-
    package_strictness %>%
    mutate(relative_strict_uniorder_function_count = round(relative_strict_uniorder_function_count, 1)) %>%
    group_by(relative_strict_uniorder_function_count) %>%
    summarize(relative_package_count = n() / total_package_count) %>%
    ungroup() %>%
    ggplot(aes(x = relative_strict_uniorder_function_count, y = relative_package_count)) +
    geom_col() + coord_flip() +
    scale_x_continuous(labels = relative_labels) +
    scale_y_continuous(sec.axis = sec_axis(~ .* total_package_count, labels=count_labels),
                       labels = relative_labels) +
    labs(x = "", y = "")

save_local_graph(package_strictness_graph, "package_strictness.pdf", width = 12, height = 3)
```

A function is strict if it evaluates all of its arguments. In the corpus, out of a total of `r format_as_absolute(CountFun)` non-zero argument functions, `r format_as_percentage(PercStrictFun)` are strict. Figure above gives a histogram of function strictness ratios per package. The majority of packages mostly contain strict functions.  The packages that are less than 75\% strict account for only `r format_as_absolute(CountStrPackLess)` packages (`r format_as_percentage(PercStrPackLess)` of all packages) and `r format_as_absolute(CountStrFunLess)` functions (`r format_as_percentage(PercStrFunLess)` of all functions).

# Parameter Strictness (Fig. 7)

The strictness of the `r format_as_absolute(CountParameter)` individual parameters positions can
also be measured. We distinguish three categories: *Always* is for
parameters that are always evaluated, *Never* is for parameters that are
never evaluated, and *Sometimes* is for parameters that we have observed
being evaluated in some calls and not in others. Figure below
illustrates the distribution of parameters across the categories.

```{r, eval=TRUE, echo=FALSE}
total_parameter_count <-
    formal_parameter_count_by_use %>%
    pull(parameter_count) %>%
    sum()

parameter_lookup_class_graph <-
    formal_parameter_count_by_usage_class %>%
    filter(parameter_use == "Lookup") %>%
    ggplot(aes(x = parameter_class, y = relative_parameter_count)) +
    scale_y_continuous(sec.axis = sec_axis(~ . * total_parameter_count, labels = count_labels),
                       labels = relative_labels) +
    geom_col() + 
    labs(x = "", y = "") + 
    coord_flip()

save_local_graph(parameter_lookup_class_graph, "parameter_lookup_class.pdf", height=2,width=12)

PercNever <-
    formal_parameter_count_by_usage_class %>% 
    filter(parameter_use=="Lookup",parameter_class=="Never") %>%
    pull(relative_parameter_count)

PercSometimes <- 
    formal_parameter_count_by_usage_class %>% 
    filter(parameter_use=="Lookup",parameter_class=="Sometimes") %>%
    pull(relative_parameter_count)

generate_macro("PercNever", PercNever)

generate_macro("PercSometimes", PercSometimes)
```
There are `r format_as_percentage(PercNever)` parameters are never evaluated, and `r format_as_percentage(PercSometimes)` parameters are inconsistently evaluated.
A strict function has all of its parameters always evaluated.

# Computation Reuse (Fig. 8)

One advantage of call-by-need is that it avoids multiple evaluations of the
same expression. How often is this used in practice? Figure below
shows, for each individual promise, how often its value was accessed.

```{r, eval = TRUE, echo = FALSE}
total_count <-  sum(argument_promise_count_by_value_lookup_count$promise_count)

argument_promise_count_by_value_lookup_count_graph <-
    argument_promise_count_by_value_lookup_count %>%
    filter(value_lookup_count %in% c(0:6, ">=7")) %>%
    ggplot(aes(value_lookup_count, relative_promise_count)) +
    geom_col() +
    coord_flip() +
    scale_y_continuous(sec.axis = sec_axis(~ . * total_count,labels = count_labels),
                       labels = relative_labels) +
    labs(x = "", y = "") + 
    scale_x_discrete(limits = c(0:6, ">=7"))

save_local_graph(argument_promise_count_by_value_lookup_count_graph, "value_lookup.pdf", width=12, height=3)

len <- nrow(argument_promise_count_by_value_lookup_count)

PercTwo <- (argument_promise_count_by_value_lookup_count%>%pull(relative_promise_count))[3]
PercTwo <- if(is.na(PercTwo)) 0 else PercTwo
PercThree <- (argument_promise_count_by_value_lookup_count%>%pull(relative_promise_count))[4]
PercThree <- if(is.na(PercThree)) 0 else PercThree
PercThousand <- sum((argument_promise_count_by_value_lookup_count%>%pull(relative_promise_count))[1001:len])
PercThousand <- if(is.na(PercThousand)) 0 else PercThousand
generate_macro("PercTwo", PercTwo)
generate_macro("PercThree", PercThree)
generate_macro("PercThousand", PercThousand)
```

The majority of promises are used once. `r format_as_percentage(PercTwo)` of promises are accessed twice,
and `r format_as_percentage(PercThree)` are accessed three times. 
`r format_as_percentage(PercThousand)` promises are accessed over 1,000 times.

# Promise Evaluation (Fig. 9)

Another advantage of laziness is that could help avoid unnecessary
computation. But do programmers avail themselves of this feature?  One way
to test this hypothesis is to compare the execution time of arguments that
are strict (promises passed to *Always* parameters) to those of lazy
arguments (promises passed to *Sometimes* parameters) and see if the
lazy arguments are more computationally intensive in any way.

Figure below shows the probability density of
promises at different running times. Promises running for less than a
millisecond are ignored.

```{r, eval=TRUE, echo=FALSE}
## convert evaluation time to seconds from milliseconds
## so that the scale can format it to appropriate time units
execution_time_graph <-
    execution_times %>%
    mutate(execution_time = log10(execution_time / 1000)) %>%
    ggplot(aes(lookup_class, execution_time,
               color = lookup_class,
               weight = relative_argument_count)) +
    geom_violin() + 
    coord_flip() + 
    theme(legend.position = "none") +
    scale_y_continuous(labels = function(x) { time_labels(10 ^ x) }) +
    labs(x = "", y = "")

save_local_graph(execution_time_graph, "argument_evaluation_time.pdf",width=12,height=4)
```
While there is a difference in the profiles, the promises that are passed to Sometimes arguments tend to be faster to evaluate. But given the large amount of promises being observed it is difficult to be conclusive.

# Function Return Type (Fig. 10)

Laziness makes it possible to compute over infinite data structures.  While
R does not provide any libraries with such data structures, it is
conceivable that some programmers may have created some in their code. 
One can use promises together with environments or
closures to create unbounded data. How often does that occur? One thing that
we can measure is *escaped promises*, the promises that outlive the
function their are passed into.

```{r,eval=TRUE,echo=FALSE}
call_by_escape_and_type <- 
    call_by_escape_and_type %>%
    rename(type=return_value_type) %>%
    filter(!(type %in% c("Unbound", "...", "Builtin", "Bytecode",
                         "Expression", "Externalptr", "Special", "Unknown (Jumped)",
                         "Pairlist", "Missing", "Weakref", "Promise", "S4", "Raw", "Complex")))
type_order <-
    call_by_escape_and_type %>%
    filter(category == "Escaped") %>%
    arrange(desc(relative_call_count)) %>%
    pull(type)

>>>>

```{r echo = TRUE, warning=FALSE, message=FALSE}
Sys.setenv("DISPLAY"=":0")

@@

bar <- function(widths, total, labels) {
    options(device = cairo_pdf)
    ## If you want to draw arbitrary rectangles, use geom_tile() or geom_rect()
    xmin <- cumsum(c(0, head(widths, -1)))
    df <- data.frame(
        xmin = xmin,
        xmax = cumsum(widths),
        ymin = -0.5,
        ymax = 0.5,
        fill = factor(1:length(widths)),
        label = labels,
        xmid = widths/2.0 + xmin,
        ymid = 0
    )

    plot <-
        df %>%
        ggplot() +
        geom_rect(aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax, fill = fill)) +
        geom_text(aes(label = label, x = xmid, y = ymid), colour = "white") +
        scale_y_continuous() +
        scale_x_continuous(breaks = cumsum(widths),
                           labels = relative_labels,
                           sec.axis = sec_axis(trans = ~ . * total,
                                               breaks = cumsum(widths),
                                               labels = count_labels)) +
        theme(axis.text.y = element_blank(),
              axis.ticks = element_blank(),
              panel.grid.minor = element_blank(),
              panel.grid.major = element_blank(),
              legend.position = "none") +
        labs(x = "", y = "")
plot
}

```

```{r echo = TRUE, warning=FALSE, message=FALSE}
promise_alive_gc_cycles <- read_table("argument_promise_alive_gc_cycle_distribution")
```

# Promise Garbage Collection

```{r, eval = TRUE, echo = FALSE}
total_argument_promise_count <- sum(promise_alive_gc_cycles$promise_count)

datatable(promise_alive_gc_cycles)

zero_gc_cycle_promises <- 
    promise_alive_gc_cycles %>%
    filter(alive_gc_cycle == 0)

if(nrow(zero_gc_cycle_promises) == 0) {
    zero_gc_cycle_promises <- tibble(promise_count = 0, relative_promise_count = 0.0)
}

one_gc_cycle_promises <- 
    promise_alive_gc_cycles %>%
    filter(alive_gc_cycle == 1)

plus_one_gc_cycle_promises <- 
    promise_alive_gc_cycles %>%
    filter(alive_gc_cycle > 1) %>%
    summarize(promise_count = sum(promise_count)) %>%
    mutate(relative_promise_count = promise_count / total_argument_promise_count)

promise_alive_gc_cycles_graph <-
    promise_alive_gc_cycles %>%
    filter(alive_gc_cycle <= 5) %>%
    ggplot(aes(x = alive_gc_cycle, y = relative_promise_count)) +
    geom_col() +
    scale_x_discrete(limits = 0:5) +
    scale_y_continuous(sec.axis = sec_axis(~ .* total_argument_promise_count, labels = count_labels),
                       labels = relative_labels) +
    coord_flip() +
    labs(x = "", y = "")

save_local_graph(promise_alive_gc_cycles_graph, "promise_alive_gc_cycles_graph.pdf", width = 12, height = 3)

widths <- c(zero_gc_cycle_promises$promise_count, one_gc_cycle_promises$promise_count, plus_one_gc_cycle_promises$promise_count)

print(widths)
bar(widths, total_argument_promise_count, c("0", "1", "> 1")) %>%
save_local_graph("promise_gc_graph.pdf", width = 12, height = 1)
#datatable()
#
```
